---
title: "foram_intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{foram_intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BPER)
```


Load a nx8 (8 column, n row(s)) data object, with columns ordered as: age, d11B, d11B2s, d11Bspec, Mg/Ca, Mg/Ca2s, d18O, d18O2s. 
The column order matters. Indicate modern species' d11B vital effects to be applied for d11Bspec. Options currently include: 'Grub' 
(G. ruber), 'Tsac' (T. sacculifer), 'Ouni' (O. universa), 'custom' (specify values below in 'priors_in_foram.R'), or 'borate'.

Run the 'load_foram' function on the input data. 
```{r}
x <- data.frame(c(55956, 55956, 55954, 55932, 55926, 55924, 55901, 55899), c(14.61, 15.46, NA, 14.49, 14.11, NA, 14.46, 13.94), 
                c(0.23, 0.22, NA, 0.34, 0.33, NA, 0.21, 0.29), c("Tsac", "Grub", NA, "Grub", "Tsac", NA, "Grub", "Tsac"), 
                c(3.26, NA, 3.33, 5.04, NA, 4.83, NA, NA), c(0.1, NA, 0.1, 0.15, NA, 0.14, NA, NA), 
                c(NA, NA, NA, -2.14, NA, -1.98, -2.19, -1.99), c(NA, NA, NA, 0.16, NA, 0.16, 0.16, 0.16))

load_foram <- load_foram(foram_data = x)
```


Take the returned list from 'load_foram' and use it for 'load_proxy' argument in 'age_index' function. Specify: the 'age_units'
used in the data set ('kyr' or 'Myr'), the time step type ('step_type'): 'regular' for evenly spaced intervals or 'every' for a time step
for every unique age in the data set. If 'regular' is specified, specify the step interval ('step_int'). 
```{r}
age_index <- age_index(load_proxy = load_foram, age_units = 'kyr', step_type = 'regular', step_int = 10)
```


Use 'clean_foram' function on the return list from 'age_index' to compile and structure observational data for the MCMC inversion.
```{r}
clean_obs <- clean_foram(age_index = age_index)
```


Revise and load user accessible parameters for the MCMC inversion. Access the package data file 'priors_in_foram.R'. This will open 
a '.R' script file in an editor window (should default to RStudio if using RStudio). This script contains the default values for 
a number of parameters, including prior distributions. Look through the script and documentation to determine which values to 
adjust. A list and description of editable parameters is also provided in the help documentation of the function 'parms_foram_adj'. 
```{r}
file.edit(system.file("extdata", "parms_in_foram.R", package = "BPER"))
```


Any of the values listed in the above '.R' script can be revised using 'parms_foram_adjust' function. There are two arguments in this 
function, for 'parms2change' provide a vector of character strings for the parameters you would like to adjust from default values.
For 'change_values' argument, provide a vector of numeric values which correspond to the parameter names in 'parms2change'. To 
demonstrate, we will adjust the temperature and d11Bsw prior distribution parameters.
```{r}
parm_names <- c("tempC.m", "tempC.sd", "d11Bsw.m", "d11Bsw.sd")
parm_values <- c(23, 4, 38.35, 0.75)

parms_foram_adj <- parms_foram_adj(parms2change = parm_names, change_values = parm_values)
```


Load in the adjusted parameters using the 'parms_forams_adj' argument OR do not include this argument to maintain default parameter 
values. Include the 'age_index' return object from the 'age_index' function in arguments. Specify which 2nd carbonate chem variable 
to use in CO2 calcs. Type ?priors_forams for list of options. Specify the type of prior 'cc2ndparm.pt' to use for the 2nd carb chem 
variable, either as an autocorrelated time series model with prior distribution defined at time step 1 ('t1') or as a time series 
with the prior distribution defined at each time step ('ts'). If 'ts' is selected, the user must include time series data object for 
'cc2ndparmTS' argument consisting of 3 columns: age, mean value for 2nd carb chem variable, and sd for 2nd carb chem variable. This 
record should span the data interval, but values do not need to be included for every time step as the record will be linearly 
interpolated to create a time-continuous record. 'priors_foram' can take several minutes to run.
```{r}

dic_pri <- data.frame(c(56200, 55934, 55928, 55919, 55900, 55883), c(2.320, 2.328, 2.405, 2.440, 2.481, 2.500), 
                      c(0.15, 0.15, 0.15, 0.15, 0.15, 0.15)) 

priors_foram <- priors_foram(parms_forams_adj = parms_forams_adj, age_index = age_index, cc2ndparm.vt = 'dic', cc2ndparm.pt = 'ts', 
                             cc2ndparmTS = dic_pri)
```


Run the MCMC inversion by supplying in the 'clean_foram' and 'priors_foram' return objects. Specify a vector of character strings 
indicating the parameters you would like saved in the MCMC output ('save.parms' argument). Also indicate the number of iterations
('n.iter'), the number of chains ('.n.chains'), the number of burn-in iterations ('n.burnin') and the amount of thinning (n.thin;
i.e., saves every 'n'th iteration). The model can take several minutes to initialize and ~minutes per 1000 iterations to run. 

A list is returned with MCMC data object 'jout' as the first object. It is recommended that you take a look at the statistical
parameters 'Rhat' and 'n.eff' in 'inv_out$jout$BUGSoutput$summary' to make sure they are acceptable.
```{r}
inv_out <- run_inversion(clean_obs = clean_obs, priors_foram = priors_foram,
                          save.parms = c("sal", "tempC", "xca", "xmg", "xso4", "d11Bsw", "d18Osw", "d18Osw.local","pco2", "dic", "pH", "press"),
                          n.iter = 10000, n.chains = 3, n.burnin = 3000, n.thin = 1)


View(inv_out$jout$BUGSoutput$summary)
```


It is recommended that chain convergence is confirmed using trace plots (i.e., built in 'trace_plot' function). Specify
the parameter of interest and the number of randomly selected trace plots for that parameter (n >1 for time-dependent parameters).
```{r}
trace_plot(inv_out = inv_out, parm = "pco2", n = 3)
```


Use the post_plot function to generate a time series plot of the posterior distributions for a time-dependent parameter. For 'type'
argument, the user can select either CI to plot the 95% credible interval, or "draws" for 500 random draws of time series posteriors.
```{r}
post_plot(inv_out = inv_out, parm = pco2, type = "CI")
```


Use the post_plot_ind function to generate a density plot for a parameter of interest. For time-dependent parameters, the time step 
to be plotted must be specified. You can also turn on/off the median line in the plot. 
```{r}
post_plot_ind(inv_out = inv_out, parm = "pco2", tstep_age = 55956, show.median = TRUE)
```


Other marine carbonate chemistry parameters can be calculated from any 2 marine carbonate chemistry parameters using 'carbchem' 
function. This can be performed on the 'inv_out' object to quantify an additional carbonate chemistry variable.
```{r}
pH <- inv_out$jout$BUGSoutput$median$pH 
ccparm2.vec <- inv_out$jout$BUGSoutput$median$pco2
temp <- inv_out$jout$BUGSoutput$median$tempC
sal <- inv_out$jout$BUGSoutput$median$sal
press <- rep_len(inv_out$jout$BUGSoutput$median$press, length.out = as.integer(length(sal)))
xca <- inv_out$jout$BUGSoutput$median$xca
xmg <- inv_out$jout$BUGSoutput$median$xmg
xso4 <- inv_out$jout$BUGSoutput$median$xso4

carbchem(ccparm2 = "pco2", ccout = "alk", pH = pH, ccparm2.vec = ccparm2.vec, temp = temp, sal = sal, press = press,
                     xca = xca, xmg = xmg, xso4 = xso4)
```
